Predicting computational reproducibility of data analysis pipelines \\in large population studies using collaborative filtering

### 2018-07-25 paper.tex
Updates in introduction and II.A

\begin{abstract}
With the rise of the reproducibility crisis in several scientific 
domains, evaluating the computational reproducibility of data analysis 
pipelines has become a critical issue. It is, however, a cumbersome 
process for analyses that involve data from large populations of 
subjects, due to their computational and storage requirements. 
We present a method to predict the computational 
reproducibility of data analysis pipelines in large population studies. 
We formulate the problem as collaborative filtering, with 
constraints on the construction of the training set. We propose 6 
different strategies to build the training set, and we evaluate them on 2 
datasets, a synthetic one modeling a population with a growing number 
of subject types, and a real one obtained with neuroinformatics 
pipelines. Results show that one sampling method, ``Random File Numbers 
(Uniform)" is able to predict computational reproducibility in 
pipelines with a good accuracy. We also analyze the relevance of 
including file and subject biases in the collaborative filtering model. 
We conclude that the proposed method is able to speed-up reproducibility 
evaluations substantially, with a reduced accuracy loss.
\end{abstract}

@@

Computational reproducibility, the property through which
computational results can be recomputed over time and
space~\cite{peng2011reproducible}, has become a critical component of scientific
methodology with the rise of the reproducibility crisis in several 
domains~\cite{baker2016there}. Among the factors hampering 
computational reproducibility, infrastructural characteristics such as 
the operating system or multithreading play an important role. In neuroinformatics, our 
primary field of interest, several studies have shown the effect of the 
operating system on computational 
results~\cite{gronenschild2012effects, glatard2015reproducibility}. 
However, conducting such reproducibility studies at scale is cumbersome 
due to the computational and storage requirements of data analysis pipelines.

Neuroinformatics pipelines are generally iterated on data coming from 10 to 
1,000 subjects, possibly with subtle input parameter variations to 
adjust specific data acquisition conditions. Subject data often capture 
anatomical and functional characteristics of their brain, for instance through
Magnetic Resonance Imaging (MRI) and Electroencephalography (EEG). The number of files 
associated with a subject may vary, and these files may also be of 
different sizes. Typical data processing times range from 15 minutes to 
15 hours, with input ranging from 100~MB to 15~GB per subject, and 
outputs ranging from 1~GB to 500~GB. Such variations impact the 
reproducibility of a given pipeline across subjects.

Our goal is to predict the outcome of reproducibility evaluations in a 
large population of subjects, from a reduced set of pipeline 
executions. More precisely, we aim at predicting whether a particular 
file produced by an analysis pipeline will be identical across 
execution conditions, or if it will contain reproducibility 
\emph{errors}. We approach the problem from the point of view of 
collaborative filtering, inspired by previous examples that applied 
this method outside of its initial application domain, for instance 
~\cite{feng2013efficient}. The main issue, and originality, of our 
problem lies in the fact that the training set cannot be arbitrarily 
sampled from the utility matrix. Instead, the training set has to 
respect the time constraints that stem from the order in which files 
are produced during pipeline executions.

This paper makes the following contributions:
\begin{enumerate}
\item We model reproducibility evaluations in data processing pipelines as a collaborative filtering problem.
\item We propose strategies to sample the training set under time constraints.
\item We evaluate and compare our sampling strategies on synthetic and real datasets.
\end{enumerate}
The problem formulation, collaborative filtering technique used, and 
proposed sampling strategies for the training set are presented in 
Section~\ref{sec:methods}. The datasets are described in 
Section~\ref{sec:datasets}, and experimental results are reported in 
Section~\ref{sec:results}.
The paper ends with a discussion about \todo{x} and a conclusion.

\todo{Complete the introduction and improve the writing}

\newpage

@@

The pipeline to be evaluated is represented by a matrix $U$ of size 
$N_f \times N_s$ where the $N_s$ columns represent data coming from 
different subjects and the $N_f$ rows represent the files produced by 
the pipeline execution. While subjects are not ordered, files are, for 
instance from their last modification time in a sequential execution. 
We assume that the order on files is consistent across subjects. $U_{i, 
j}$ measures the reproducibility of file $i$ produced during the 
processing of subject $j$ in two conditions, for instance two different 
operating systems. In our experiments, we restrict $U_{i,j}$ to be 
boolean, but our methods can be applied for real values as well.

@@

Collaborative filtering is a technique to predict unknown values of a 
matrix called "utility matrix" from the known ones. Traditionally, the 
matrix represents the ratings of items, represented in columns, by 
users, represented in rows. Ratings might be explicit, when users 
provide ratings through a dedicated system, or implicit, when users' 
behaviors are tracked and analyzed to estimate their preferences. An 
overview of collaborative filtering is provided 
in~\cite{leskovec2014mining}. In our context, the utility matrix is the
pipeline matrix $U$ described previously.

@@

@@

% Generalizability

@@

% Real values, we expect the best sampling method to remain the same.
We recommend not to include biases in the collaborative filtering optimization
to solve this problem. Even though biases provide a slight accuracy improvement 
when datasets are strongly biased (constant lines in the matrix), they can also
be very detrimental for more complex datasets such as the synthetic one used here.

% Other possible examples
The method described in this paper and applied to reproducibility 
studies could be used to predict the outcome of other deterministic, 
hidden, parameterized, time-dependent processes, for instance markers 
of chronicle disease activity.

We consider the problem of predicting the values of a variable 
controlled by a hidden, parameterized, time-dependent, deterministic 
process. We assume that the variable values are observed only for a 
limited number of time points and process parameters, and we aim at 
predicting the values for the remaining time points and parameters.

This problem description maps directly to the evaluation of 
computational reproducibility in scientific pipelines: the process is 
the evaluated pipeline, it is hidden since we consider the pipeline as a black box (we 
don't inspect its source code), it is parametrized as it behaves 
differently on different input datasets and parameters, it is 
time-dependent as the value of the variable of interest varies 
throughout the course of the pipeline execution. The variable of 
interest is whether the pipeline outputs, most likely the files it 
produces, are identical between execution conditions.

@@

RFNTS and RFNTL could be used when the matrix is more  uncertain at the beginning or end of the pipeline.

RS and RFNU tends to become the same for high training ratios.

>>>>

\begin{abstract}
With the rise of the reproducibility crisis, evaluating the 
computational reproducibility of data analysis pipelines has become a 
critical issue. It is, however, a cumbersome 
process for analyses that involve data from large populations of 
subjects, due to their computational and storage requirements. 
We present a method to predict the computational 
reproducibility of data analysis pipelines in large population studies. 
We formulate the problem as a collaborative filtering process, with 
constraints on the construction of the training set. We propose 6 
different strategies to build the training set, which we evaluate on 2 
datasets, a synthetic one modeling a population with a growing number 
of subject types, and a real one obtained with neuroinformatics 
pipelines. Results show that one sampling method, ``Random File Numbers 
(Uniform)" is able to predict computational reproducibility with a good 
accuracy. We also analyze the relevance of including file and subject 
biases in the collaborative filtering model. We conclude that the 
proposed method is able to speed-up reproducibility 
evaluations substantially, with a reduced accuracy loss.
\end{abstract}

@@

Computational reproducibility, the ability to recompute data analyses 
over time and 
space~\cite{peng2011reproducible}, has become a critical component of 
scientific methodology as many researchers acknowledge the existence of 
a reproducibility crisis~\cite{baker2016there}. Among other factors, 
such as data and code sharing, infrastructural characteristics 
play an important role in making experiments reproducible. For instance, in 
neuroinformatics, our primary field of interest, several studies have 
shown the effect of the 
operating system on computational 
results~\cite{gronenschild2012effects, glatard2015reproducibility}.
However, conducting such reproducibility studies at scale is cumbersome 
due to the computational and storage requirements of data analysis 
pipelines.

Neuroinformatics pipelines are generally iterated on data coming from 10 to 
1,000 subjects, possibly with subtle input parameter variations to 
adjust specific data acquisition conditions. Subject data often capture 
anatomical and functional characteristics of their brain, for instance through
Magnetic Resonance Imaging (MRI) or Electroencephalography (EEG). The number of files 
associated with a subject may vary, and these files may also be of 
different sizes. Typical data processing times range from 15 minutes to 
15 hours per subject, with input ranging from 100~MB to 15~GB per subject, and 
outputs ranging from 1~GB to 500~GB.

The reproducibility of a given pipeline may vary across subjects, for 
instance due to different pipeline branches being executed depending on 
the nature, cardinality or content of the input data. For instance, in 
the public database released by the Human Connectome 
Project~\cite{van2013wu}, subjects may have one or two anatomical images of 
each modality; when two images are 
present, they are aligned together and averaged. The presence of 
artifacts, for instance due to high motion during the acquisition, may 
also trigger corrections that are not otherwise required. Some branches 
of the pipeline may also be executed only for specific acquisition 
parameters, for instance the intensity of the magnetic field. To 
capture all these variations, reproducibility evaluations need to be 
conducted on many subjects, which is unwieldy.

Our goal is to predict the outcome of reproducibility evaluations in a 
large population of subjects, from a reduced set of pipeline 
executions. More precisely, we aim at predicting whether a particular 
file produced by an analysis pipeline will be identical across 
execution conditions, or if it will contain reproducibility 
\emph{errors}. We approach the problem from the point of view of 
collaborative filtering, inspired by previous examples that applied 
this method outside of its initial application domain, for instance 
~\cite{feng2013efficient}. The main issue, and originality, of our 
problem lies in the fact that the training set cannot be arbitrarily 
sampled from the utility matrix. Instead, the training set has to 
respect the time constraints imposed by the order in which files 
are produced during pipeline executions.

This paper makes the following contributions:
\begin{enumerate}
\item We model reproducibility evaluations in data processing pipelines as a collaborative filtering problem.
\item We propose strategies to sample the training set under time constraints.
\item We evaluate and compare our sampling strategies on synthetic and real datasets.
\end{enumerate}
The problem formulation, collaborative filtering technique used, and 
proposed sampling strategies for the training set are presented in 
Section~\ref{sec:methods}. The datasets are described in 
Section~\ref{sec:datasets}, and experimental results are reported in 
Section~\ref{sec:results}.
The paper ends with a discussion about the impact, limitations and 
generalizability of our results.

@@

The pipeline to be evaluated is represented by a matrix $U$ of size 
$N_f \times N_s$, where the $N_s$ columns represent data coming from 
different subjects, and the $N_f$ rows represent the files produced by 
the pipeline execution. While subjects are not ordered, files are, for 
instance from their last modification time in a sequential execution, 
and we assume that this order is consistent across subjects. $U_{i, j}$ 
measures the reproducibility of file $i$ produced during the processing 
of subject $j$ in two conditions, for instance two different 
operating systems. In our experiments, we restrict $U_{i,j}$ to be 
boolean, but our methods can be applied for real values as well.

@@

Collaborative filtering is a technique to predict unknown values of a 
matrix called ``utility matrix" from the known ones. Traditionally, the 
matrix represents the ratings of items, represented in columns, by 
users, represented in rows. Ratings might be explicit, when users 
provide ratings through a dedicated system, or implicit, when users' 
behaviors are tracked and analyzed to estimate their preferences. An 
overview of collaborative filtering is provided 
in~\cite{leskovec2014mining}. In our context, the utility matrix is the
pipeline matrix $U$ described previously.

@@

In any case, we recommend not to include biases in the collaborative filtering 
optimization to solve this problem. Even though biases provide a slight 
accuracy improvement when datasets are strongly biased (constant lines 
in the matrix), they can also be very detrimental for more complex 
datasets such as the synthetic one used here.

@@

@@

% Other possible examples
The method described in this paper and applied to reproducibility 
studies could be used to predict the other variables controlled by 
hidden, deterministic, hidden, parameterized, time-dependent processes, 
for instance markers of chronicle disease activity. In the 
reproducibility example studied in this paper, the variable of interest 
is the reproducibility of files produced by the processing of a given 
subjects, the process is the evaluated pipeline, it is hidden since we 
consider the pipeline as a black box (we don't inspect its source 
code), it is parametrized as it behaves differently on different input 
datasets and parameters, it is time-dependent as the value of the 
variable of interest varies throughout the course of the pipeline 
execution.

@@
